False Positive Probabilities for all Kepler Objects of Interest: \\ \nvalnew\ newly validated planets and \nfpnew\ likely false positives

### 2016-04-20 document/ms.tex
re-submitted to apj
something

\setlength{\tabcolsep}{2pt}

@@

\newcommand{\nlongfp}{530}

@@

%\author{Timothy D. Morton}
%\affil{Department of Astrophysical Sciences, Princeton University}
\author{Timothy D. Morton\altaffilmark{1}, Stephen T. Bryson\altaffilmark{2}, Jeffrey L. Coughlin\altaffilmark{2,3}, Jason F. Rowe\altaffilmark{4}, Ganesh Ravichandran\altaffilmark{5}, Erik A. Petigura\altaffilmark{6}, 
Michael R. Haas\altaffilmark{2}, and Natalie M. Batalha\altaffilmark{2}}

@@

\altaffiltext{1}{Department of Astrophysical Sciences, Princeton University, 4 Ivy Lane, Princeton, NJ 08544, USA, tdm@astro.princeton.edu}
\altaffiltext{2}{NASA Ames Research Center, M/S 244-30, Moffett Field, CA 94035, USA}
\altaffiltext{3}{SETI Institute, 189 Bernardo Ave, Mountain View, CA 94043, USA}
\altaffiltext{4}{D\'epartement de Physique, Universit\'e de Montr\'eal, Montr\'eal, QC, Canada, H3T 1J4}
\altaffiltext{5}{Columbia University}
\altaffiltext{6}{California Institute of Technology}

@@

The basic idea of \vespa\ is to assign probabilities to different
hypotheses that might describe a transiting planet
candidate signal.  If $\{H_i\}$ is the set of all considered
hypotheses, the probability for any given model $i$ is
\begin{equation}
  \eqlabel{prob}
  \mathrm{Pr}\left(H_i\right) = \frac{\pi_i \mathcal
    L_i}{\displaystyle \sum_j \pi_j \mathcal L_j},
\end{equation}
where $\pi_i$ is the ``hypothesis prior'' and $\mathcal L_i$ is the
``hypothesis likelihood''\footnote{This factor is more widely known as
  the ``Bayesian evidence'' or ``marginalized likelihood'';
  \citet{Morton:2014b} argues for the term ``hypothesis likelihood,''
  as it can be clarifying to think of it that way.}
The prior represents how intrinsically probable the hypothesized
scenario is to exist, and the likelihood represents how closely the
shape of the observed transit signal matches with the expected shape
of a signal produced by the hypothesis.

@@

We note that a built-in weakness of model selection is that it assumes that
the set of models being considered is comprehensive.  This could
in principle lead to a situation where one model is strongly preferred
over all other models, but even that model is a poor explanation of
the data---in this work, this could lead to improperly validated planets.  
There are two general strategies to try to address this issue.  The first 
is to somehow quantify the absolute goodness-of-fit of the models, and 
require that a validated planet pass some threshold test.  The other 
strategy, that we adopt here, is to expand the set of models to be more 
comprehensive.  In order to do this, we introduce two artificial models:
``boxy'' and ``long.''  The ``boxy'' likelihood function is a step function
at some minimum value of $T/\tau$ (zero below this threshold, and constant
above), and constant throughout the space of the other trapezoidal parameters.
Similarly, the likelihood of the ``long'' model is a step function at some
minimum threshold value of duration $T$.  These thresholds  
are set relative to the simulated trapezoidal shapes of the planet model:
the $T/\tau$ threshold is the maximum value from the simulated planet population,
and the $T$ threshold is the 99\% percentile of simulated planet
population values.  We also choose the model priors for these artificial
models to be low, reflecting that we expect only a small number of signals
to be unexplained by any of the astrophysical scenarios: the number that we
choose for each of these models is $5 \times 10^{-5}$, corresponding to an expectation
that there may be $\sim$10 such signals among the $\sim$200,000 \kepler\ targets.

@@

In order to properly interpret these results, it is necessary to
understand the range of applicability of the \vespa\ calculation.
First of all, this method selects between different specific explanations 
for the transit-like signal, and cannot
comment on whether the signal might be caused by stellar variability
or an instrumental false alarm.  Thus, \vespa\ results on low
signal-to-noise ratio (SNR) candidates that are not clearly
transit-like must be viewed with caution.  This being said, the reason 
for including the artificial 
``boxy'' and ``long'' models in the model selection calculation
 (\sect{methods:fpp}) is to flag signals that do not 
fit well with any of the astrophysical eclipse models---in fact, the ``long''
model is preferred by \nlongfp\ KOIs that are already dispositioned
as FALSE POSITIVE.

Additionally, an important constraint used in the FPP calculation is the
allowed sky area inside which a blended false positive may live.  As
described in \sect{data}, this value is taken to be three times the
uncertainty on the fitted centroid position from the pixel-level data.
However, many KOIs have already been identified to be blended binary
false positives displaced from the target star.  Some of these are
found by detecting significant centroid offsets in the pixel-level
data---in these cases, \vespa\ treating the confusion radius simply as
the uncertainty in the centroid position will clearly give a
misleading result.  Other displaced false positives have been
identified as originating from displaced stars by finding KOIs with
matching periods and epochs \citep{Coughlin:2014}, and often the
``parents'' of these signals are outside the pixel masks, and so
unable to be detected via centroid-measuring methods.  In these cases
as well, the \vespa\ assumptions break down, and the FPP calculations
will not be valid.

@@

A much more independent and powerful test data set has recently become
available in the work of \citet{Santerne:2015}, which presented the
results of a long-term RV-monitoring campaign targeting 129 \kepler\
giant-planet candidates.  Of these, they confirm 45 to be planets and
identify 48 as eclipsing binaries, 15 as contaminating EBs (CEBs), and
3 as brown dwarfs.  They are unable to determine the nature of the
remaining 18.  These results imply a relatively high FP rate among
giant-planet candidates, possibly near 50\%, which sounds potentially
concerning, although we reiterate that KOIs are not ruled FALSE 
POSITIVES based on their transit depth or inferred size alone.
Additionally, when we look at the \vespa\ results on this sample, we
see very good agreement between our results and the RV-
detected ``ground truth'' (\Tab{santerne}).  Notably, confirmed
planets show a mean FPP of about 10\% and a median of much less than 1\%,
while confirmed EBs (CEBs) show a mean FPP of 75\% (78\%) and a median
of 97\% (99\%).  The three brown dwarfs also show low FPPs, which is
understandable because \vespa\ doesn't pretend to predict anything
about the mass of the companion, and BDs are essentially the same size
as giant planets.  It is also instructive to investigate the four
cases where \vespa\ computes high FPPs ($>$0.5) for confirmed planets.
Three of these (KOI-377.01, KOI-1426.02 and KOI-1474.01) have
significant TTVs, and one is a grazing eclipse (KOI-614.01).
KOI-1474.01 is also on a highly eccentric orbit \citep{Dawson:2012},
in addition to its TTVs, which also contributes to its high FPP.
Overall, this comparison powerfully demonstrates the reliability of
the \vespa\ calculation, showing that even in a population of
candidates that include many false positives, it is able to
effectively identify which are true planets and which are not.

@@

We also briefly discuss another case of individual interest.
KOI-6705.01, a 0.99\,d signal around a mid M-dwarf star, was
identified  as a KOI of possible interest by \citet{Gaidos:2015}.
After significant follow-up observations and detailed full-frame image
analysis, they concluded that the signal was most likely due to a
charge-transfer  effect from a 1.99\,d EB located on the same CCD
column.  For this KOI, \vespa\ calculates an FPP of 1, with the 
``long'' model preferred by far.

probabilities of 0.86 to be a BEB, 0.13 to be a BEB with twice the
period (the true scenario uncovered by the detective work of
\citet{Gaidos:2015}), and 0.01 to be an HEB.  This is an excellent
example of what was discussed in \sect{PEMs}: that even effects like
column anomalies that were not identified by the \kepler\ team as
ephemeris matches will typically be identified as false positives by 
\vespa.

@@

\begin{figure*}[p]
\begin{center}
\includegraphics[width=7in]{figures/fppalpha.pdf}
\end{center}
\caption{Periods and radii of KOIs with CANDIDATE and CONFIRMED disposition.  
  Blue circles have previously been identified as CONFIRMED.  Candidates are orange circles, 
  shaded by false positive probability, with 
  a transparent circle representing a high FPP. 
  \figlabel{rpcand}}
\end{figure*}

>>>>

@@

@@

%\author{Timothy D. Morton}
%\affil{Department of Astrophysical Sciences, Princeton University}
\author{Timothy D. Morton\altaffilmark{1}, Stephen T. Bryson\altaffilmark{2}, Jeffrey L. Coughlin\altaffilmark{2,3}, Jason F. Rowe\altaffilmark{4},Michael R. Haas\altaffilmark{2}, and Natalie M. Batalha\altaffilmark{2}}

@@

\altaffiltext{1}{Department of Astrophysical Sciences, Princeton University, 4 Ivy Lane, Princeton, NJ 08544, USA, tdm@astro.princeton.edu}
\altaffiltext{2}{NASA Ames Research Center, M/S 244-30, Moffett Field, CA 94035, USA}
\altaffiltext{3}{SETI Institute, 189 Bernardo Ave, Mountain View, CA 94043, USA}
\altaffiltext{4}{D\'epartement de Physique, Universit\'e de Montr\'eal, Montr\'eal, QC, Canada, H3T 1J4}

@@

The basic idea of \vespa\ is to assign probabilities to different
astrophysical hypotheses that might describe a transiting planet
candidate signal.  If $\{H_i\}$ is the set of all considered
hypotheses, the probability for any given model $i$ is
\begin{equation}
  \eqlabel{prob}
  \mathrm{Pr}\left(H_i\right) = \frac{\pi_i \mathcal
    L_i}{\displaystyle \sum_j \pi_j \mathcal L_j},
\end{equation}
where $\pi_i$ is the ``hypothesis prior'' and $\mathcal L_i$ is the
``hypothesis likelihood''\footnote{This factor is more widely known as
  the ``Bayesian evidence'' or ``marginalized likelihood'';
  \citet{Morton:2014b} argues for the term ``hypothesis likelihood,''
  as it can be clarifying to think of it that way.}
The prior represents how intrinsically probable the hypothesized
scenario is to exist, and the likelihood represents how closely the
shape of the observed transit signal matches with the expected shape
of a signal produced by the hypothesis.

@@

@@

In order to properly interpret these results, it is necessary to
understand the range of applicability of the \vespa\ calculation.
First of all, this method selects between different \emph{eclipse
  configuration} explanations for the transit-like signal, and cannot
comment on whether the signal might be caused by stellar variability
or an instrumental false alarm.  Thus, \vespa\ results on low
signal-to-noise ratio (SNR) candidates that are not clearly
transit-like must be viewed with caution.

However, while the FPP value does only give the \emph{relative}
probability between the false positive and planet scenarios, there is
a way to get a sense of whether \emph{all} of the models might be bad
fits to the data (and thus get a hint that perhaps the signal is due
to stellar variability or a false alarm).  Recall that FPP is
calculated as
\begin{equation}
\eqlabel{fpp}
{\rm FPP} = \frac{\displaystyle\sum_{i \neq {\rm Pl}}
  L_i}{\displaystyle \sum_i L_i}
\end{equation}
where $L_i$ = $\mathcal L_i \pi_i$ and $i$ runs over all the models,
with ``Pl'' representing the planet scenario.  If, for example, the
shape of the transit is such that \emph{none} of the models gives any
kind of decent fit to its trapezoidal shape, then $L_{\rm tot} =
\sum_i L_i$ will be very small.  It appears that a value of $10^{-3}$
is a reasonable threshold below which to consider a particular
calculation unreliable.  To set this threshold we inspected the
handful of systems that have low FPPs and $L_{\rm tot} < 10^{-3}$
(some go as low as $10^{-10}$ or lower), and find that these very
often have unrealistically long durations for a transit signal.  Such
a case could easily give the planet hypothesis the \emph{largest}
likelihood (resulting in a low FPP), not because it is intrinsically
likely, but because all the other scenarios are even less likely.

Another important constraint used in the FPP calculation is the
allowed sky area inside which a blended false positive may live.  As
described in \sect{data}, this value is taken to be three times the
uncertainty on the fitted centroid position from the pixel-level data.
However, many KOIs have already been identified to be blended binary
false positives displaced from the target star.  Some of these are
found by detecting significant centroid offsets in the pixel-level
data---in these cases, \vespa\ treating the confusion radius simply as
the uncertainty in the centroid position will clearly give a
misleading result.  Other displaced false positives have been
identified as originating from displaced stars by finding KOIs with
matching periods and epochs \citep{Coughlin:2014}, and often the
``parents'' of these signals are outside the pixel masks, and so
unable to be detected via centroid-measuring methods.  In these cases
as well, the \vespa\ assumptions break down, and the FPP calculations
will not be valid.

@@

A much more independent and powerful test data set has recently become
available in the work of \citet{Santerne:2015}, which presented the
results of a long-term RV-monitoring campaign targeting 129 \kepler\
giant-planet candidates.  Of these, they confirm 45 to be planets and
identify 48 as eclipsing binaries, 15 as contaminating EBs (CEBs), and
3 as brown dwarfs.  They are unable to determine the nature of the
remaining 18.  These results imply a relatively high FP rate among
giant-planet candidates, possibly near 50\%, which sounds potentially
concerning, although we reiterate that KOIs are not ruled FALSE 
POSITIVES based on their transit depth or inferred size alone.
Additionally, when we look at the \vespa\ results on this sample, we
see very good agreement between our results and the RV-
detected ``ground truth'' (\Tab{santerne}).  Notably, confirmed
planets show a mean FPP of about 9\% and a median of only 0.02\%,
while confirmed EBs (CEBs) show a mean FPP of 78\% (83\%) and a median
of 95\% (99.5\%).  The three brown dwarfs also show low FPPs, which is
understandable because \vespa\ doesn't pretend to predict anything
about the mass of the companion, and BDs are essentially the same size
as giant planets.  It is also instructive to investigate the four
cases where \vespa\ computes high FPPs ($>$0.3) for confirmed planets.
Three of these (KOI-377.01, KOI-1426.02 and KOI-1474.01) have
significant TTVs, and one is a grazing eclipse (KOI-614.01).
KOI-1474.01 is also on a highly eccentric orbit \citep{Dawson:2012},
in addition to its TTVs, which also contributes to its high FPP.
Overall, this comparison powerfully demonstrates the reliability of
the \vespa\ calculation, showing that even in a population of
candidates that include many false positives, it is able to
effectively identify which are true planets and which are not.

@@

We also briefly discuss another case of individual interest.
KOI-6705.01, a 0.99\,d signal around a mid M-dwarf star, was
identified  as a KOI of possible interest by \citet{Gaidos:2015}.
After significant follow-up observations and detailed full-frame image
analysis, they concluded that the signal was most likely due to a
charge-transfer  effect from a 1.99\,d EB located on the same CCD
column.  For this KOI, \vespa\ calculates an FPP of 1, with
probabilities of 0.86 to be a BEB, 0.13 to be a BEB with twice the
period (the true scenario uncovered by the detective work of
\citet{Gaidos:2015}), and 0.01 to be an HEB.  This is an excellent
example of what was discussed in \sect{PEMs}: that even effects like
column anomalies that were not identified by the \kepler\ team as
ephemeris matches will typically be identified as false positives by 
\vespa.

@@

\begin{figure*}[p]
\begin{center}
\includegraphics[width=7in]{figures/RP_cand.pdf}
\end{center}
\caption{Periods and radii of KOIs with CANDIDATE disposition, grouped
  by FPP value.  Green circles are most likely to be planets, red X's
  are most likely to be false positives, and blue plusses have
  intermediate FPPs.  Grey dots are KOIs not identified as likely 
  false positives for which the
  \vespa\ calculation is considered less reliable.  
  \figlabel{rpcand}}
\end{figure*}
